Studying measurement invariance in factor analysis models via stochastic processes

### 2011-03-23 www/paper.Rnw
Paper moved to "www" directory and turned into .Rnw file.
All results in the manuscript can now be generated dynamically using Sweave().
Hence, no static PDF graphics (except the model diagram) are required
and these will be generated on the fly.

Also renamed .bib and added lavaan reference.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@56 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
\title{Studying measurement invariance in factor analysis models
  via stochastic processes}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{Wichita State University}{Universit\"{a}t Innsbruck}

@@

@@

To employ the LR test for assessing measurement invariance, model
parameters are estimated separately for a certain number of subgroups
of the data (with some parameters potentially restricted to be equal across
subgroups).\footnote{For ease of exposition, we describe
the case where there are no such parameter restrictions.  As shown in
the example and simulation below, it is
straightforward to extend all methods to the more general case.}
Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.

Analogously to the LR test, the Wald test and LM test (also known as score test)
can be employed to test the null hypothesis $H_1^*$ for a fixed threshold $\nu$.
For the Wald test, the idea is to compute the Wald statistic $W(\nu)$ as a quadratic
form in $\hat {\bm \theta}^{(A)} - \hat {\bm \theta}^{(B)}$, utilizing its estimated
covariance matrix for standardization. For the LM test, the LM statistic
$\mathit{LM}(\nu)$ is a quadratic form in ${\bm s}(\hat {\bm \theta}; x_1, \dots, x_m)$
and ${\bm s}(\hat {\bm \theta}; x_{m+1}, \dots, x_n)$. Thus, the three tests all assess
differences that should be zero under $H_0$: for the LR test the difference of maximized
likelihoods; for the Wald test, the difference of parameter estimates; and for the
LM test, the differences of likelihood scores from zero. In the LR case, the parameters
have to be estimated under both the null hypothesis and alternative.
Conversely, the Wald case requires  
only the estimates under the alternative, while the LM case requires
only the estimates under the null hypothesis.

@@

As pointed out above, specification of the asymptotic distribution under $H_0$ for the test
statistics from the previous section is straightforward: It is simply the aggregation
of the asympotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
from (\ref{eq:dmax}) converges to $\sup_t || {\bm B}^0(t) ||_{\infty}$, where
$||\cdot||_\infty$ denotes the maximum norm. Similarly, $\mathit{CvM}$
from (\ref{eq:cvm}) converges to $\int_0^1 || {\bm B}^0(t) ||_2^2 d t$, where $||\cdot||^2$
denotes the Euclidean norm. Finally, $\max \mathit{LM}$ from (\ref{eq:maxlm}) -- and analogously
the maximum Wald and LR tests -- converges to $\sup_t (t (1-t))^{-1} || {\bm B}^0(t) ||_2^2$
(which can also be interpreted as the maximum of a tied-down Bessel process, as pointed out
previously).

@@

If a measurement invariance violation is detected by any of the tests,
one may want to incorporate it into the model to account for it. The
procedure for doing this  typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
discussed above often prove helpful in determining a suitable parameterization.
In particular, one approach that is often employed in practice
involves adoptions of a model
with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
single threshold case). In the multiple threshold case, their location can
be determined by maximizing the corresponding segmented log-likelihood over all
possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
algorithm to this task). For the single threshold case, this reduces to
maximizing the segmented log-likelihood
\begin{equation} \label{eq:seglik}
  \ell(\hat {\bm \theta}^{({A)}}; x_1, \dots, x_m) ~+~ \ell(\hat {\bm \theta}^{({B)}}; x_{m+1}, \dots, x_n)
\end{equation}
over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR statistic
from (\ref{eq:lr}) (with some minimal subgroup size typically imposed).

@@

\begin{figure}
\caption{Path diagram representing the base factor analysis model used for
  the example and simulations.  To induce measurement invariance
  violations, a seventh observed variable (student age) determines the
values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
$\lambda_{31}$).}
\label{fig:famod}
\includegraphics[height=5in]{famod.pdf}
%\input{figs/path}
\end{figure}

@@

@@

@@

Figure~\ref{fig:lrlm} 
compares the max-LM statistic (solid line) to the max-LR
statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
\eqref{eq:h0}.  The critical values for these two
tests are identical, hence the single horizontal line.
The figure
shows that the two statistics are very similar to one another, with
both maxima at the dotted vertical line.  This
is generally to be expected, because the two tests are asymptotically
equivalent.  The 
max-LR statistic cannot be obtained from the empirical fluctuation
process, however, so the factor analysis model must be refitted before
and after each of the possible threshold values $\nu$.

@@

\begin{figure}
\caption{Three test statistics of \eqref{eq:exh0} and \eqref{eq:h0}, 
  based on the example involving measurement invariance with respect
  to student age.  Solid, horizontal lines represent critical values
  at $\alpha=.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\includegraphics[height=9in]{gefp_3_19.pdf}
\end{figure}

\begin{figure}
\caption{A comparison of the max-LM (solid line) and max-LR (dashed line) test
  statistics for \eqref{eq:h0}.  The solid horizontal line
  corresponds to the critical value at $\alpha=.05$ while the dotted vertical
  line highlights the threshold at which the test statistics are maximized.}
\label{fig:lrlm}
\includegraphics[height=5in]{lmlr_19.pdf}
\end{figure}

\begin{figure}
\caption{Cumulative score processes for each verbal factor
  loadings, with critical values stemming from the double-max test. 
  The solid, horizontal lines
  correspond to the critical value at $\alpha=.05$.}
\label{fig:lambdas}
\includegraphics[height=5in]{gefp_lambdas.pdf}
\end{figure}

@@

% Simulations demonstrating power of test(s) at various
% sample sizes, types of invariance violations, ?
In this section, we conduct a simulation designed to
examine the tests' power and Type I error rates in a context
similar to the example.
We examine the power and error rates of three tests: the double-max test, the
Cram\'{e}r-von Mises test, and the max-LM test.  We also compare 
tests involving only the parameters that changed with tests of all model
parameters.  The example implied that the former tests had more
power, and the simulations provide more detail on the extent to which
the tests' power levels differ.
Finally, we also examine the tests' power across
various magnitudes of measurement invariance violations.

\subsection{Method}
Data were generated in a similar way to that of 
the example section.
The data included verbal scales that violated measurement invariance
with respect to students' ages, with students 13--15 years of age
having smaller factor loadings than students 16--18 years of age.  
Simulation data were generated from the same general factor analysis
model (see Figure~\ref{fig:famod}), but sample size and magnitude
of measurement 
invariance violation were manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and fifteen magnitudes of violations.  These violations
involved the younger students' values of $\{\lambda_{11},
\lambda_{21}, \lambda_{31}\}$ deviating from the older students'
values by $d$ 
times the parameters' asymptotic standard errors, with $d = 0, 0.25,
0.5, \dots, 3.5$.  The 0-standard error condition was
used to study Type I error rate.

@@

@@

\begin{figure}
\caption{Simulated power for three test statistics across three sample
  sizes, two subsets of tested parameters, and measurement invariance
  violations of 0--3.5 standard 
  errors (scaled by $n$).  Circles reflect the double-max test,
  triangles reflect the 
  max-LM test, and plus signs reflect the Cram\'{e}r-von Mises test.
  Numbers above each panel reflect the sample size and number of
  parameters being tested, respectively.}
\label{fig:simres}
\includegraphics[height=9in]{simres.pdf}
\end{figure}

\begin{table*}
\caption{Simulated power for three test statistics
  across three sample sizes, fifteen magnitudes of measurement
  invariance violations, and two subsets of tested
  parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
  test; Max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
\label{tab:simres}
\begin{center}
\begin{tabular}{lllllllllll}
  \hline
   & & & \multicolumn{8}{l}{Violation Magnitude (SE)} \\
  $n$ &  npars &  Statistic    &  0 & 0.5 &    1 &  1.5
       &    2 &  2.5  &   3 &  3.5 \\ 
  \hline
100 & 3 & DM & 3.1 & 4.1 & 8.5 & 16.1 & 27.2 & 43.2 & 57.1 & 67.4 \\ 
   &  & CvM & 4.9 & 6.6 & 13.1 & 26.4 & 45.0 & 65.9 & 80.4 & 89.2 \\ 
   &  & Max-LM & 4.9 & 6.8 & 11.4 & 20.6 & 34.8 & 54.8 & 71.4 & 82.9 \\ 
   & 19 & DM & 2.4 & 3.0 & 4.3 & 5.9 & 9.5 & 15.9 & 23.8 & 31.5 \\ 
   &  & CvM & 4.6 & 4.4 & 7.0 & 11.1 & 17.1 & 27.1 & 38.1 & 49.2 \\ 
   &  & Max-LM & 7.0 & 7.1 & 9.0 & 11.2 & 16.2 & 21.4 & 31.0 & 39.7 \\ 
  200 & 3 & DM & 3.7 & 5.0 & 9.0 & 17.8 & 32.0 & 49.3 & 67.1 & 84.0 \\ 
   &  & CvM & 4.9 & 6.5 & 13.9 & 26.2 & 48.2 & 68.1 & 85.5 & 95.1 \\ 
   &  & Max-LM & 5.5 & 6.1 & 11.1 & 20.9 & 39.1 & 58.6 & 78.5 & 92.0 \\ 
   & 19 & DM & 3.8 & 3.4 & 4.6 & 7.1 & 13.1 & 22.0 & 34.7 & 52.5 \\ 
   &  & CvM & 4.6 & 4.9 & 7.3 & 11.8 & 20.1 & 31.7 & 45.8 & 64.0 \\ 
   &  & Max-LM & 6.0 & 6.1 & 8.2 & 11.3 & 16.3 & 24.4 & 37.6 & 54.3 \\ 
  500 & 3 & DM & 3.9 & 5.6 & 9.7 & 19.5 & 34.3 & 52.3 & 71.9 & 87.0 \\ 
   &  & CvM & 4.8 & 7.3 & 12.9 & 26.9 & 47.3 & 69.1 & 86.7 & 94.9 \\ 
   &  & Max-LM & 5.5 & 6.7 & 11.4 & 21.4 & 38.5 & 60.1 & 81.4 & 93.2 \\ 
   & 19 & DM & 3.7 & 4.4 & 5.3 & 8.7 & 14.6 & 25.6 & 41.6 & 60.9 \\ 
   &  & CvM & 4.8 & 5.3 & 7.1 & 11.9 & 20.6 & 32.3 & 50.3 & 68.8 \\ 
   &  & Max-LM & 5.6 & 6.2 & 7.5 & 10.4 & 15.9 & 25.8 & 41.6 & 60.8 \\ 
   \hline
\end{tabular}
\end{center}
\end{table*}

@@

\subsection{Use in Practice}
% Simultaneous test of Meredith's various types of invariance
The proposed tests give researchers a new set of tools for studying 
measurement invariance.  For example, they give researchers the
flexibility to: (1)~simultaneously test all model parameters, yielding
results relevant to many types of measurement
invariance (see, e.g., \citeNP{Mer93}), or (2)~test a single subset of
model parameters, potentially leading to improved power to detect a
single type of measurement invariance.  The traditional
steps have involved hypothesizing a specific type of invariance and
then testing for it via a LR test, but this is
unnecessary under the proposed framework.

@@

All results were obtained using the {R}~system for statistical computing \cite{R2010},
version~2.11.1, employing the add-on packages {OpenMx}~0.9.1-1421 \cite{BokNea11} for
fitting of the factor analysis models and {strucchange}~1.4.2 \cite{ZeiLei02,Zei06}
for evaluating the parameter instability tests. {R} and {strucchange} are
freely available under the General Public License~2 from the
Comprehensive {R} Archive Network at \url{http://CRAN.R-project.org/}
while {OpenMx} is available under the
Apache License~2.0 from \url{http://openmx.psyc.virginia.edu/}.
R code for replication of our results is
available at \url{http://semtools.R-Forge.R-project.org/}.

\bibliography{list}

>>>>

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

\title{Studying measurement invariance in factor analysis models
  via stochastic processes}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{Wichita State University}{Universit\"{a}t Innsbruck}

@@

<<preliminaries>>=
library("OpenMx")
library("lavaan")
library("strucchange")
library("mvtnorm")
library("lattice")
source("mz.R")
source("sim.R")
@

@@

To employ the LR~test for assessing measurement invariance, model
parameters are estimated separately for a certain number of subgroups
of the data (with some parameters potentially restricted to be equal across
subgroups).\footnote{For ease of exposition, we describe
the case where there are no such parameter restrictions.  As shown in
the example and simulation below, it is
straightforward to extend all methods to the more general case.}
Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR~test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.

Analogously to the LR~test, the Wald test and LM test (also known as score test)
can be employed to test the null hypothesis $H_1^*$ for a fixed threshold $\nu$.
For the Wald test, the idea is to compute the Wald statistic $W(\nu)$ as a quadratic
form in $\hat {\bm \theta}^{(A)} - \hat {\bm \theta}^{(B)}$, utilizing its estimated
covariance matrix for standardization. For the LM test, the LM statistic
$\mathit{LM}(\nu)$ is a quadratic form in ${\bm s}(\hat {\bm \theta}; x_1, \dots, x_m)$
and ${\bm s}(\hat {\bm \theta}; x_{m+1}, \dots, x_n)$. Thus, the three tests all assess
differences that should be zero under $H_0$: for the LR~test the difference of maximized
likelihoods; for the Wald test, the difference of parameter estimates; and for the
LM test, the differences of likelihood scores from zero. In the LR~case, the parameters
have to be estimated under both the null hypothesis and alternative.
Conversely, the Wald case requires  
only the estimates under the alternative, while the LM case requires
only the estimates under the null hypothesis.

@@

As pointed out above, specification of the asymptotic distribution under $H_0$ for the test
statistics from the previous section is straightforward: It is simply the aggregation
of the asympotic process ${\bm B}^0(t)$ \cite{HjoKon02,ZeiHor07}. Thus, $\mathit{DM}$
from (\ref{eq:dmax}) converges to $\sup_t || {\bm B}^0(t) ||_{\infty}$, where
$||\cdot||_\infty$ denotes the maximum norm. Similarly, $\mathit{CvM}$
from (\ref{eq:cvm}) converges to $\int_0^1 || {\bm B}^0(t) ||_2^2 d t$, where $||\cdot||^2$
denotes the Euclidean norm. Finally, $\max \mathit{LM}$ from (\ref{eq:maxlm}) -- and analogously
the maximum Wald and LR~tests -- converges to $\sup_t (t (1-t))^{-1} || {\bm B}^0(t) ||_2^2$
(which can also be interpreted as the maximum of a tied-down Bessel process, as pointed out
previously).

@@

If a measurement invariance violation is detected by any of the tests,
one may want to incorporate it into the model to account for it. The
procedure for doing this  typically depends on the type of violation ${\bm \theta}(V)$, and the visualizations
discussed above often prove helpful in determining a suitable parameterization.
In particular, one approach that is often employed in practice
involves adoptions of a model
with one (or more) threshold(s) in all parameters (i.e., (\ref{eq:h1*}) for the
single threshold case). In the multiple threshold case, their location can
be determined by maximizing the corresponding segmented log-likelihood over all
possible combinations of thresholds (\citeNP{ZeiSha10} adapt a dynamic programming
algorithm to this task). For the single threshold case, this reduces to
maximizing the segmented log-likelihood
\begin{equation} \label{eq:seglik}
  \ell(\hat {\bm \theta}^{({A)}}; x_1, \dots, x_m) ~+~ \ell(\hat {\bm \theta}^{({B)}}; x_{m+1}, \dots, x_n)
\end{equation}
over all values of $m$ corresponding to possible thresholds $\nu$ (such that $v_m \le \nu$ and
$v_{m+1} > \nu$). As pointed out previously, this is equivalent to maximizing the LR~statistic
from (\ref{eq:lr}) (with some minimal subgroup size typically imposed).

@@

\begin{figure}
\caption{Path diagram representing the base factor analysis model used for
  the example and simulations.  To induce measurement invariance
  violations, a seventh observed variable (student age) determines the
values of the verbal factor loadings ($\lambda_{11}$, $\lambda_{21}$,
$\lambda_{31}$).}
\label{fig:famod}
\includegraphics[height=5in]{famod.pdf}
\end{figure}

@@

<<dgp>>=
set.seed(1090)
d <- dgp(200, 2)
@

@@

<<mz_mod>>=
if(file.exists("mz_mod.rda")) {
  load("mz_mod.rda")
} else {
  ## full sample model
  mz0 <- mzfit(d)
  ## subsample models
  lrstat <- function(age) 2 * as.vector(logLik(mzfit(d[d$age <= age,])) + logLik(mzfit(d[d$age > age,])) - logLik(mz0))
  mz.age <- sort(d$age[d$age >= quantile(d$age, 0.1) & d$age <= quantile(d$age, 0.9)])
  mz.lrstat <- sapply(mz.age, lrstat)
  save(mz0, mz.age, mz.lrstat, file = "mz_mod.rda")
}
@

@@

Figure~\ref{fig:lrlm} 
compares the max-LM statistic (solid line) to the max-LR
statistic (dashed line) from \eqref{eq:maxlr}, as applied to testing
\eqref{eq:h0}.  The critical values for these two
tests are identical, hence the single horizontal line.
The figure
shows that the two statistics are very similar to one another, with
both maxima at the dotted vertical line.  This
is generally to be expected, because the two tests are asymptotically
equivalent.  The 
max-LR~statistic cannot be obtained from the empirical fluctuation
process, however, so the factor analysis model must be refitted before
and after each of the possible threshold values $\nu$.

@@

\begin{figure}
\caption{Three test statistics of \eqref{eq:exh0} and \eqref{eq:h0}, 
  based on the example involving measurement invariance with respect
  to student age.  Solid, horizontal lines represent critical values
  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\setkeys{Gin}{width=\textwidth}
<<gefp-3-19, fig=TRUE, height=8, width=6>>=
## empirical fluctuation processes (3 vs all 19 pars)
gefp_3_info <- gefp(mz0, fit = NULL, order.by = d$age, parm = 1:3,
  vcov = info.mzfit, sandwich = FALSE)
gefp_19_info <- gefp(mz0, fit = NULL, order.by = d$age,
  vcov = info.mzfit, sandwich = FALSE)

## visualization
par(mfcol = c(3, 2))
plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
plot(gefp_3_info,  functional = supLM(0.1), main = "max-LM, k* = 3", xlab = "Age")
plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
plot(gefp_19_info, functional = supLM(0.1), main = "max-LM, k* = 19", xlab = "Age")
@
\end{figure}

\begin{figure}
\caption{A comparison of the max-LM (solid line) and max-LR (dashed line) test
  statistics for \eqref{eq:h0}.  The solid horizontal line
  corresponds to the critical value at $\alpha=.05$ while the dotted vertical
  line highlights the threshold at which the test statistics are maximized.}
\label{fig:lrlm}
\setkeys{Gin}{width=\textwidth}
<<lm-lr-19, fig=TRUE, height=5, width=6>>=
plot(gefp_19_info, functional = supLM(0.1),
  xlab = "Age", ylab = "LR and LM statistics", main = "", ylim = c(0, 50))
legend("topleft", c("LR", "LM"), lty = c(2, 1), bty = "n")
lines(mz.age, mz.lrstat, lty = 2)
abline(v = mz.age[which.max(mz.lrstat)], lty = 3)
@
\end{figure}

\begin{figure}
\caption{Cumulative score processes for each verbal factor
  loadings, with critical values stemming from the double-max test. 
  The solid, horizontal lines
  correspond to the critical value at $\alpha=.05$.}
\label{fig:lambdas}
\setkeys{Gin}{width=\textwidth}
<<gefp-lambdas, fig=TRUE, height=8, width=6>>=
plot(gefp_3_info,  functional = maxBB, aggregate = FALSE, ylim = c(-2, 2),
  main = "", xlab = "Age", ylab = expression(lambda[11], lambda[21], lambda[31]))
@
\end{figure}

@@

In this section, we conduct a simulation designed to
examine the tests' power and Type I error rates in a context
similar to the example.
We examine the power and error rates of three tests: the double-max test, the
Cram\'{e}r-von Mises test, and the max-LM test.  We also compare 
tests involving only the parameters that changed with tests of all model
parameters.  The example implied that the former tests had more
power, and the simulations provide more detail on the extent to which
the tests' power levels differ.
Finally, we also examine the tests' power across
various magnitudes of measurement invariance violations.

\subsection{Method}
Data were generated in a similar way to that of 
the example section.
The data included verbal scales that violated measurement invariance
with respect to students' ages, with students 13--15 years of age
having smaller factor loadings than students 16--18 years of age.  
Simulation data were generated from the same general factor analysis
model (see Figure~\ref{fig:famod}), but sample size and magnitude
of measurement 
invariance violation were manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and fifteen magnitudes of violations.  These violations
involved the younger students' values of $\{\lambda_{11},
\lambda_{21}, \lambda_{31}\}$ deviating from the older students'
values by $d$ 
times the parameters' asymptotic standard errors, with $d = 0, 0.25,
0.5, \dots, 4$.  The 0-standard error condition was
used to study Type I error rate.

@@

<<mz_sim>>=
if(file.exists("mz_sim.rda")) {
  load("mz_sim.rda")
} else {
  ## seed for replication
  RNGkind(kind = "default", normal.kind = "default")
  set.seed(1090)
  ## run simulation
  mz_sim <- simulation()
  save(mz_sim, file = "mz_sim.rda")
}

## select only subset
mz_sim_sub <- subset(mz_sim, pars %in% c(3, 19))
mz_sim_sub$pars <- factor(mz_sim_sub$pars)
## labeling
levels(mz_sim_sub$pars) <- paste("k* =", levels(mz_sim_sub$pars))
levels(mz_sim_sub$nobs) <- paste("n =", levels(mz_sim_sub$nobs))
@

@@

\begin{figure}
\caption{Simulated power for three test statistics across three sample
  sizes, two subsets of tested parameters, and measurement invariance
  violations of 0--3.5 standard 
  errors (scaled by $n$).  Circles reflect the double-max test,
  triangles reflect the 
  max-LM test, and plus signs reflect the Cram\'{e}r-von Mises test.
  Numbers above each panel reflect the sample size and number of
  parameters being tested, respectively.}
\label{fig:simres}
\setkeys{Gin}{width=\textwidth}
<<mzsim-xyplot, fig=TRUE, height=7, width=7>>=
## visualization
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub, type = "b",
  xlab = "Violation Magnitude", ylab = "Power", ylim = c(0, 1)))
@
\end{figure}

\begin{table*}
\caption{Simulated power for three test statistics
  across three sample sizes, fifteen magnitudes of measurement
  invariance violations, and two subsets of tested
  parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
  test; max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
\label{tab:simres}
\begin{center}
\begin{tabular}{lrlrrrrrrrrr}
  \hline
   & & & \multicolumn{9}{l}{Violation Magnitude (SE)} \\
<<mz_sim-table, results=tex>>=
## labeling
levels(mz_sim_sub$pars) <- gsub("k* = ", "", levels(mz_sim_sub$pars), fixed = TRUE)
levels(mz_sim_sub$nobs) <- gsub( "n = ", "", levels(mz_sim_sub$nobs), fixed = TRUE)
levels(mz_sim_sub$test) <- gsub("supLM", "max-LM", levels(mz_sim_sub$test), fixed = TRUE)

## table
mz_tab <- round(ftable(100 * xtabs(power ~ nobs + pars + test + diff,
  data = mz_sim_sub, subset = diff %in% c(seq(0, 4, by = 0.5))),
  col.vars = "diff"), digits = 1)
mz_tab <- format(mz_tab, quote = FALSE)[-2, -4]
mz_tab[1,] <- c("$n$", "$k^*$", "Statistic", format(seq(0, 4, by = 0.5)))
mz_tab <- paste(apply(mz_tab, 1, paste, collapse = " & "), "\\\\")
mz_tab[c(1, length(mz_tab))] <- paste(mz_tab[c(1, length(mz_tab))], "\\hline")
writeLines(mz_tab)
@

\end{tabular}
\end{center}
\end{table*}

@@

\subsection{Use in Practice}
% Simultaneous test of Meredith's various types of invariance
The proposed tests give researchers a new set of tools for studying 
measurement invariance.  For example, they give researchers the
flexibility to: (1)~simultaneously test all model parameters, yielding
results relevant to many types of measurement
invariance (see, e.g., \citeNP{Mer93}), or (2)~test a single subset of
model parameters, potentially leading to improved power to detect a
single type of measurement invariance.  The traditional
steps have involved hypothesizing a specific type of invariance and
then testing for it via a LR~test, but this is
unnecessary under the proposed framework.

@@

All results were obtained using the {R}~system for statistical computing \cite{R11},
version~\Sexpr{paste(R.Version()[6:7], collapse = ".")}, employing the add-on packages
{lavaan}~\Sexpr{packageDescription("lavaan")["Version"]} \cite{lavaan11} and
{OpenMx}~\Sexpr{packageDescription("OpenMx")["Version"]} \cite{BokNea11} for fitting of the factor analysis models and
{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \cite{ZeiLei02,Zei06}
for evaluating the parameter instability tests. {R}~and the packages {lavaan} and {strucchange} are
freely available under the General Public License~2 from the
Comprehensive {R} Archive Network at \url{http://CRAN.R-project.org/}
while {OpenMx} is available under the
Apache License~2.0 from \url{http://openmx.psyc.virginia.edu/}.
{R}~code for replication of our results is
available at \url{http://semtools.R-Forge.R-project.org/}.

\bibliography{refs}
